Network Concurrency Control Done In Band

### 2015-05-30 inband-sync.tex
intro

\begin{abstract}
Control planes of forthcoming Software-Defined Networks (SDNs)
will be \emph{distributed}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules operating on the data plane should be physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, %a distributed control plane also introduces
%new challenges,
in order to guarantee \emph{consistency} of network operation, actions
performed on the data plane by different controllers may need to be \emph{synchronized},  which is a
 nontrivial task.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
In this paper, we propose a synchronization
framework for control planes  based on the well-known concept of
\emph{transactions}.
%
A transaction aggregates a sequence
of control messages and specific synchronization primitives,
providing all-or-nothing semantics.
We argue that a powerful transactional interface should and can be implemented \emph{inband}, using 
standard OpenFlow commands on the data plane configuration.
%Indeed, recent OpenFlow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
%In particular, we present a
%transactional interface which supports arbitrary atomic
%transactions
%over the switch configuration space:
%Our transactions
%can include complex notions of conflicts (beyond flow overlap checks),
%as well as computations and multiple controller interactions.
We discuss how to use our synchronization framework to implement
 consistent policy composition and fault-tolerant control-plane applications.
\end{abstract}

@@

By consolidating and outsourcing the control over the dataplane switches to a logically
centralized controller, Software-Defined Networks (SDNs)
simplify network management and facilitate faster innovations:
the (software) control plane can evolve independently from the
(hardware) data plane.
Moreover, OpenFlow, the standard SDN protocol today, introduces many generalizations,
in terms of traffic engineering, definition of flows, as well as in-band network functionalities,
by relying on a simple match-action paradigm which allows us to define
forwarding rules based not only on Layer-2, but also Layer-3 and Layer-4 (and possibly beyond).

While the perspective of \emph{logically centralized} control plane
offered by SDN is intuitive and attractive,
there is a wide consensus that
the control plane should be  \emph{physically distributed}.
First, in order to provide high availability,
%as well as a minimal degree of
%fault-tolerance,
%[[PK fault-tolerance is necessary to provide hight availability?]]
controllers should be redundant~\cite{onix,onos,elasticon}: a failure 
of one controller can be masked by other controllers.
Second, it has also been proposed to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication-intensive
%control plane
data plane
events close to their origin~\cite{devoflow,kandoo}.
Third, larger SDNs are likely to be operated by multiple administrators~\cite{stn} or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently~\cite{participatory}.

@@

Today, we do not have a good understanding yet of how to realize
such distributed control planes. The problem is essentially a
distributed systems
one: Multiple controllers may simultaneously try to
install conflicting updates and we want to resolve these conflicts
\emph{consistently} (no undesired behavior is observed on the data
plane) and \emph{efficiently} (no undesired delays are imposed on the
control application).
\emph{Synchronizing}
the distributed controllers and manipulating the network state \emph{consistently}
are non-trivial tasks.~\cite{sharon}

@@

The concurrent computing literature offers
a wide range of synchronization abstractions.
E.g., the popular \emph{transactional
  memory} abstraction~\cite{tm-book} provides a collection of
concurrent processes with the ability of aggregating sequences of
shared-memory operations in atomic \emph{transactions} with
all-or-nothing semantics.

@@

\noindent\textbf{Contributions}
This paper applies the principle of atomicity in concurrent computing
to distributed SDN control planes.
%
In particular, we propose synchronization constructs which
allow a controller to represent multiple configuration commands on
the data plane as an atomic transaction.
%
If none of the transaction commands \emph{conflicts}  with the current
configuration, where conflicts can be
defined in a general, application-specific manner, the transaction appears to be executed atomically.
Otherwise, the transaction is aborted in its entirety and does not
affect neither traffic nor other controllers.

We promote to implement our synchronization constructs \emph{inband},
on the OpenFlow switch.
An inband implementation allows us to efficiently alleviate the problems related to
coordinating controllers via a control plane
out-of-band network, in the presence of asynchrony and failures.
Indeed, the inherent costs  of such protocols are often considered too high, both in
terms of the necessary computability assumptions about the underlying
system~\cite{FLP85}, and the amount of communication
complexity~\cite{Lam06}.
In contrast, our \emph{in-band} solution allows the
controllers to solve fundamental agreement tasks in just one message
exchange with the data plane, tolerating asynchrony and failures of
any number of controllers.

Interestingly, our mechanisms are simple and can be implemented using the \emph{standard OpenFlow}
protocol (version 1.4). 
%
%By providing more complex and atomic transactions over the
%switch configuration
%space, several synchronization problems can be resolved.
%This paper investigates the implementation of a transactional interface
%(\`{a} la STN~\cite{stn}) modified concurrently by multiple controllers.
%\petr{"interface modified concurrently" does not sound right -> "accessed concurrently"?}
%
Concretely, in our implementation, 
as part of the data plane configuration space, 
we propose to use a shared memory storing information about  contention and
conflicts between distributed controllers.
A transaction can then contain standard OpenFlow control operations
as well as  \emph{synchronization primitives} operating on this shared
memory.
Our synchronization primitives allow the controllers to define complex notions
of conflicts that can be used for coordinating configuration updates operating 
over \emph{independent flow spaces}, but where dependencies are introduced
in the mapping of the flows to the physical infrastructure. This is
important, e.g., for load-balancing applications.

@@

We then show that, using our abstraction, the controllers can
solve several important synchronization problems.
We describe an implementation of the fundamental
\emph{compare-and-set} (CAS) primitive,
which in turn can be used to solve consensus and
implement a generic replicated control plane service~\cite{Her91} in a consistent and
fault-tolerant way: a mainstream building block in
modern system design.
We also discuss a simple CAS-based concurrent policy-update mechanism,
and also use our synchronization abstraction to provide
the missing link for the read-modify-write object
postulated in~\cite{stn}. \petr{We do not seem to discuss the latter?}

@@

More generally, CAS can be used to solve the fundamental
problem of fault-tolerant \emph{consensus}~\cite{FLP85}
\hide{%[[PK, feels like we do not really need this level of detail here
 in which the controllers
need to agree on one of their private inputs, despite failures of an
arbitrary number of controllers.
Indeed, each controller simply performs, within a transaction, a CAS operation to replace the default value
in a memory location at the meta-configuration of a data plane switch.
If the transaction commits, the controller considers itself
the \emph{winner} and decides on its input,
if it aborts (the controller fails in the CAS), it remains to read the
memory to get the input of the winner and decide on it.
}
and allows us
to implement any generic replicated state machine~\cite{Her91}
in the control plane,  in the \emph{wait-free} manner, i.e., tolerating
asynchrony and failure of any number of controllers.
As such, CAS can also be used to implement STN~\cite{stn},
by providing the missing synchronization object.

>>>>

\begin{abstract}
Control planes of forthcoming Software-Defined Networks (SDNs)
will be \emph{distributed}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules operating on the data plane should be physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, %a distributed control plane also introduces
%new challenges,
in order to guarantee \emph{consistency} of network operation, actions
performed on the data plane by different controllers may need to be \emph{synchronized},  which is a
 nontrivial task.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
In this paper, we propose a synchronization
framework for control planes  based on the well-known concept of
\emph{transactions}.
%
A transaction aggregates a sequence
of control messages and specific synchronization primitives,
providing all-or-nothing semantics.
We argue that a powerful transactional interface should and can be
implemented \emph{in-band}, using 
standard OpenFlow commands on the data plane configuration.
%Indeed, recent OpenFlow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
%In particular, we present a
%transactional interface which supports arbitrary atomic
%transactions
%over the switch configuration space:
%Our transactions
%can include complex notions of conflicts (beyond flow overlap checks),
%as well as computations and multiple controller interactions.
We discuss how to use our synchronization framework to implement
 consistent policy composition and fault-tolerant control-plane applications.
\end{abstract}

@@

By consolidating and outsourcing the control over the dataplane switches to a logically
centralized controller, Software-Defined Networks (SDNs)
simplify network management and facilitate faster innovations:
the (software) control plane can evolve independently from the
(hardware) data plane.
%Moreover, OpenFlow, the standard SDN protocol today, introduces many generalizations,
%in terms of traffic engineering, definition of flows, as well as in-band network functionalities,
%by relying on a simple match-action paradigm which allows us to define
%forwarding rules based not only on Layer-2, but also Layer-3 and Layer-4 (and possibly beyond).
%
While the perspective of \emph{logically centralized} control plane
offered by SDN is intuitive and attractive,
there is a wide consensus that
the control plane should be  \emph{physically distributed}.
First, in order to provide high availability,
%as well as a minimal degree of
%fault-tolerance,
%[[PK fault-tolerance is necessary to provide hight availability?]]
controllers should be redundant~\cite{onix,onos,elasticon}: a failure 
of one controller can be masked by other controllers.
Second, it has also been proposed to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication-intensive
%control plane
data plane
events close to their origin~\cite{devoflow,kandoo}.
Third, larger SDNs are likely to be operated by multiple administrators~\cite{stn} or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently~\cite{participatory}.

@@

Today, we do not have a good understanding yet of how to realize
such distributed control planes. The problem is essentially a
distributed-systems
one: Multiple controllers may simultaneously try to
install conflicting updates and we want to resolve these conflicts
\emph{consistently} (no undesired behavior is observed on the data
plane) and \emph{efficiently} (no undesired delays are imposed on the
control application).
\emph{Synchronizing}
the distributed controllers and manipulating the network state \emph{consistently}
are non-trivial tasks.~\cite{sharon}

@@

The concurrent computing literature offers
a wide range of synchronization abstractions.
E.g., the popular \emph{transactional
  memory} abstraction~\cite{tm-book} provides a collection of
concurrent processes with the ability of aggregating sequences of
shared-memory operations in \emph{atomic  transactions} with
all-or-nothing semantics.

@@

\hspace{1mm}
\noindent\textbf{Contributions.}
This paper applies the principle of atomicity in concurrent computing
to distributed SDN control planes.
%
In particular, we propose synchronization constructs which
allow a controller to represent multiple configuration commands on
the data plane as an atomic transaction.
%
If none of the transaction's commands \emph{conflicts}  with the current
configuration, where conflicts can be
defined in a general, application-specific manner, the transaction appears to be executed atomically.
Otherwise, the transaction is \emph{aborted} in its entirety and does not
affect neither traffic nor other controllers.

We promote to implement our synchronization constructs \emph{in-band},
on the switch.
An in-band implementation allows us to efficiently alleviate the problems related to
coordinating controllers via a control plane
out-of-band network, in the presence of asynchrony and failures.
Indeed, the inherent costs  of such protocols are often considered too high, both in
terms of the necessary computability assumptions about the underlying
system~\cite{FLP85}, and the amount of communication
complexity~\cite{Lam06}.
In contrast, our \emph{in-band} solution allows the
controllers to solve fundamental agreement tasks in just one message
exchange with the data plane, tolerating asynchrony and failures of
any number of controllers.

Interestingly, our mechanisms are simple and can be implemented using the \emph{standard OpenFlow}
protocol (version 1.4). 
%
%By providing more complex and atomic transactions over the
%switch configuration
%space, several synchronization problems can be resolved.
%This paper investigates the implementation of a transactional interface
%(\`{a} la STN~\cite{stn}) modified concurrently by multiple controllers.
%\petr{"interface modified concurrently" does not sound right -> "accessed concurrently"?}
%
Concretely, in our implementation, 
we use a part of the data-plane configuration space as a \emph{shared
  memory}  that stores information about  contention and conflicts
between the controllers.
A transaction can then contain standard OpenFlow control operations
as well as  \emph{synchronization primitives} operating on this shared
memory.
Our synchronization primitives allow the controllers to define
general notions of conflicts between configuration updates, 
not restricted to simple conflicts on the overlapping flow spaces.
In particular, we can define dependencies between mappings of
independent flows to the physical infrastructure, which is important,
e.g., for load-balancing applications.     
%which can be used even for coordinating 
%configuration updates operating 
%over \emph{independent flow spaces}, but where dependencies are introduced
%in the mapping of the flows to the physical infrastructure. This is
%important, e.g., for load-balancing applications.

@@

We then show that, using our abstraction, the controllers can
solve several important synchronization problems.
We describe an implementation of the fundamental
\emph{compare-and-set} (CAS) primitive,
which in turn can be used to solve consensus and
implement a generic replicated control plane service~\cite{Her91} in a consistent and
fault-tolerant way: a mainstream building block in
modern system design.
We also discuss a simple CAS-based concurrent policy-update mechanism,
and also use our synchronization abstraction to provide
the missing link for the read-modify-write object
postulated in~\cite{stn}. %\petr{We do not seem to discuss the latter?}

@@

More generally, CAS can be used to solve the fundamental
problem of fault-tolerant \emph{consensus}~\cite{FLP85}
\hide{%[[PK, feels like we do not really need this level of detail here
 in which the controllers
need to agree on one of their private inputs, despite failures of an
arbitrary number of controllers.
Indeed, each controller simply performs, within a transaction, a CAS operation to replace the default value
in a memory location at the meta-configuration of a data plane switch.
If the transaction commits, the controller considers itself
the \emph{winner} and decides on its input,
if it aborts (the controller fails in the CAS), it remains to read the
memory to get the input of the winner and decide on it.
}
and allows us
to implement any generic replicated state machine~\cite{Her91}
in the control plane,  in the \emph{wait-free} manner, i.e., tolerating
asynchrony and failure of any number of controllers.
As such, CAS can also be used to implement STN~\cite{stn},
by providing the postulated \emph{read-modify-write} object.
